---
title: "R Widget template - Not Configured for External Use - CC and DD Category: Air Quality-related Respiratory Illness v1"
author: "Robert Sucsy"
date: "`r Sys.Date()`"
output: html_document
params:
  state_abbreviation: "ri"
  end_date_of_interest: !r Sys.Date()
  start_date_of_interest: !r Sys.Date()-90
  API_of_rawdata_of_interest: "https://essence.syndromicsurveillance.org/nssp_essence/api/dataDetails?geography=ri&datasource=va_er&startDate=9Mar2023&medicalGroupingSystem=essencesyndromes&userId=5576&endDate=1Jun2023&percentParam=noPercent&aqtTarget=DataDetails&ccddCategory=air%20quality-related%20respiratory%20illness%20v1&geographySystem=state&detector=probrepswitch&timeResolution=daily"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

#defines CRAN repo
r = getOption("repos")
r["CRAN"] = "http://cran.us.r-project.org"
options(repos = r)

#installs Rnssp since it's not in CRAN
if (require(Rnssp) == TRUE) {
  library(Rnssp)
} else if (require(devtools) == TRUE) {
  devtools::install_github("cdcgov/Rnssp")
  library(Rnssp)
} else {
  install.packages("devtools")
  devtools::install_github("cdcgov/Rnssp")
  library(Rnssp)
}

#Installs CRAN packages
#for final make sure package lists are correct
list.of.packages <- c("tidyverse","httr","jsonlite","knitr","ggthemes","lubridate","tidycensus","ggmap","plotly","leaflet","stats","Hmisc","rpivotTable","dplyr", "janitor", "epitools")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) {install.packages(new.packages)}

lapply(list.of.packages, require, character.only = TRUE)

#Code Stopped running? Make sure password is up to date for the profile
load("~/myProfile.rda") 
```

The following is for internal use only.This is a preliminary surveillance effort into health outcomes related to the current poor air quality due to Canada wildfires.

The health outcome in particular used for aggregation is the Chief Complaint & Discharge Diagnosis Category: Air Quality-related Respiratory Illness v1. This is a combination of multiple aqm-related health outcomes; for specifics see page 7 in the link **[here](https://cste-my.sharepoint.com/personal/aaltabbaa_cste_org/_layouts/15/onedrive.aspx?id=%2Fpersonal%2Faaltabbaa%5Fcste%5Forg%2FDocuments%2FAlyaa%20Altabbaa%2FWildfire%20and%20Smoke%20Syndromic%20Surveillance%5FFINAL%2Epdf&parent=%2Fpersonal%2Faaltabbaa%5Fcste%5Forg%2FDocuments%2FAlyaa%20Altabbaa&ga=1)**.

The data is from the CDC's ESSENCE database, a syndromic surveillance databse originating from hospital records.Regarding data quality, there appears to be a lag in reporting, so it will be worth ongoing examination.
(NSSP ESSENCE: National Syndromic Surveillance Program – 
Electronic Surveillance System for the Early Notification of Community-based Epidemics)


Additional note: That bright ZIP code in SW RI is centered on Chariho middle and high schools, and otherwise has  a low population, which is likely the source of it's anomalous results

If there are further questions or if an additional tool is required please contact robert.sucsy@health.ri.gov


```{r config, include=FALSE}
geography<-params$state_abbreviation
geography<-tolower(geography)

end_date_of_interest<-params$end_date_of_interest
start_date_of_interest<-params$start_date_of_interest

end_date_of_interest<-as.Date(end_date_of_interest)
start_date_of_interest<-as.Date(start_date_of_interest)

endDate <- format(end_date_of_interest, "%d%b%Y")
startDate <- format(start_date_of_interest, "%d%b%Y")

datestring<-as.data.frame(seq(as.Date(start_date_of_interest), as.Date(end_date_of_interest), "days"))
datestring$Date = datestring$`seq(as.Date(start_date_of_interest), as.Date(end_date_of_interest), "days")`
datestring<-subset(datestring, select = Date)

load("~/myProfile.rda") 
```

```{r importing raw data, include=FALSE}

rawdata_url<-params$API_of_rawdata_of_interest
rawdata_url<-str_replace(rawdata_url,"geography=[a-z]+", paste0("geography=",geography))
rawdata_url<-str_replace(rawdata_url ,"startDate=\\d{1,2}\\w{3}\\d{4}", paste0("startDate=",startDate))
rawdata_url<-str_replace(rawdata_url ,"endDate=\\d{1,2}\\w{3}\\d{4}", paste0("endDate=",endDate))


rawdata_pull<-get_api_data(rawdata_url)
rawdata<-rawdata_pull$dataDetails

zipsums<-rawdata %>% group_by(ZipCode) %>% summarise(count=length(Date))
datesums<-rawdata %>% group_by(Date) %>% summarise(count=length(Date))
datesums$Date<-as.Date(datesums$Date, tryFormats = c("%m/%d/%Y"))

datesums<-left_join(datestring,datesums)
datesums$count<- ifelse(is.na(datesums$count), 0, datesums$count)

state_wide<-get_acs(geography = "zip code tabulation area",
                 variables = c(tot_pop = "B01001_001"),
                 state = geography,
                 year = 2019,
                 geometry = TRUE,
                 output = "wide")

countzips<-left_join(state_wide, zipsums, by = c("GEOID" = "ZipCode"))
countzips$countper10000<-round(10000*countzips$count/countzips$tot_popE, digits = 3)
```

```{r age-adjusted rate, include=FALSE}
acs5vars2021<-load_variables(2021,"acs5", cache=T)

varlist<-acs5vars2021[grep("B01001", acs5vars2021$name),] 
varlist<-varlist[1:49,1:2]
varlist$name<-paste0(varlist$name,"E")
varlist<-varlist %>% add_row(name = "GEOID", label = "GEOID")
varlist<-varlist %>% add_row(name = "NAME", label = "NAME")
varlist<-varlist %>% add_row(name = "geometry", label = "geometry")

ri_blocks<-get_acs(geography = "block group",
                   table = "B01001",
                   state = "RI",
                   year = 2021,
                   geometry = TRUE,
                   output = "wide")

ri_blocks<-ri_blocks[,c(1,2,seq(3, 101, by = 2))]

names(ri_blocks) <- varlist$label[match(names(ri_blocks), varlist$name)]

ri_blocks<-ri_blocks %>% clean_names()

ri_block_centroid<-st_centroid(ri_blocks)

joined_zctas_and_blocks<-st_join(state_wide,ri_block_centroid)

joined_zctas_and_blocks<-joined_zctas_and_blocks[,-c(2:6)]

zcta_summaries<-joined_zctas_and_blocks %>% group_by(GEOID, geometry) %>% summarise(across(everything(), list(sum), na.rm = TRUE))
```

```{r map, include = FALSE}
#leaflet plot
colorpalette <- colorNumeric("viridis", countzips$countper10000)

label_asthma <- function(GEOID, countper10000){
  str_glue("Zip code {GEOID} had {countper10000} outcomes of interest per ten thousand people")
}


plot3<-countzips %>% 
  leaflet() %>% 
  addPolygons(weight = 1,
              color = "white",
              fillColor = ~colorpalette(countper10000),
              fillOpacity = 1,
              popup = ~label_asthma(GEOID, countper10000)) %>% 
  addTiles() %>% 
  addLegend(pal = colorpalette,
            values = ~countper10000,
            title = "Count per </br>10,000 People",
            opacity = 1)
```

```{r asthma time series plots,include=FALSE}

plot4<-ggplot(datesums, aes(x= Date)) + geom_smooth(aes(y= count), method = "loess", formula = "y~x",span = 7/90) +
  theme_classic()+
  ggtitle("Count, Past 90 days")+
  geom_point(aes(y= count)) 
plot4

plot4<-ggplotly(plot4)

```

```{r text generation, echo=FALSE}
header_text<-paste0("There were ", length(rawdata$Visit_ID)," outcomes of interest ", start_date_of_interest," and ",end_date_of_interest)
```

# `r header_text`

```{r print, echo = FALSE}
plot3
plot4
rawdata_subset<-rawdata[,c("HospitalName", "ZipCode", "Region", "Sex", "Age", "c_ethnicity", "c_race", "AgeGroup")]
rpivotTable(data = rawdata_subset, rows = "AgeGroup",cols="Sex", vals = "Freq", aggregatorName = "Count", rendererName = "Table")
```


